---
title: "RegMod-Project. Research on the effect of automatic transmission on mileage per gallon (MPG)"
author: "Gerrit Versteeg"
date: "July 2nd, 2017"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## 1. Executive summary
This report investigates the influence of automatic versus manual transmission on the mileage per gallon for cars. The results indicate that for a car of average weight and an average 1/4-mile time, a manual transmission adds 2.94 miles per gallon on the average 9.62 miles per gallon of a car with an automatic transmission (95% certainty). So using manual transmission is better for fuel-efficiency.

## 2. Loading and preparing the data
First step is to load the necessary libraries and the 'mtcars' data from the library 'datasets'. We then put the data into a data frame (tibble) and make the numeric variable 'am' a factor variable.

```{r data preparation, echo=TRUE, message=FALSE, warning=FALSE}
library("GGally", quietly=TRUE, warn.conflicts = FALSE)    ## for ggpairs
library("dplyr", quietly=TRUE, warn.conflicts = FALSE)     ## for data prep
library("ggplot2", quietly=TRUE, warn.conflicts = FALSE)   ## for the graphs
library("datasets", quietly=TRUE, warn.conflicts = FALSE)  ## for the data iself
DF <- tbl_df(mtcars)               ## read data into tibble
DF$am <- as.factor(DF$am)          ## set am from num to factor
attach(DF)                         ## to reference column-names
```

## 2. Exploratory Data Analysis
There are no missing values (NULL, NA's) for any of the variables in mtcars, so nothing needs to be imputed via some imputing strategy. First let's have a look at all variables and their correlations to get an impression of the dataset by looking at the ggpairs plot (see appendix, figure 1). The variables that have the largest correlation with our independent variable "am" are in descending order: gear (0.79), drat (0.71), wt (-0.69), disp (-0.59), cyl (-0.52). The remaining variables appear to be further specifications of the type of engine. Also there is a 0.6 correlation between am and mpg and the plot suggests an increase in mpg for cars with a manual transmission.

## 3. Choosing the model
A normal linear model seems in order, because the outcomes (mpg) are not binary, Bernoulli nor binomial and there are no unbounded count data, rates or proportions. Modelling a first fit with mpg as outcome and am as sole predictor, yields a promising result.

```{r firstfit, ECHO = TRUE}
fit1 <- lm(mpg~am)                         ## fitting just am
summary(fit1)$coef
```

The expected mpg for cars with automatic transmission is 17.15 mpg, while the expected mpg for cars with manual transmission is 24.39 (an additional 7.24 miles per US gallon). The relationship appears to be strongly significant (2.85e-04). BUT the adjusted R-squared is only 0.3385, therefor only 34% of the variation in the dataset is explained through this first simple model.

So it is necessary to determine which further variables to take into account for linear regression and to check for possible confounders that have an impact on the relationship between 'am' and 'mpg'. A factor analysis, like PCA, is possible but dangerous because I lack the knowledge how to do that properly. Therefor we revisit the variables relying on common sense and general understanding of cars (the subject matter at hand).

Looking at the various variables in mtcars, we see that a lot of lesser correlated (with am) variables have to do with the type of engine (horsepower, no of cylinders, no of carburators, V/S-engine, displacement). To decrease the number of potential variables, consider the variable qsec (1/4 mile time, acceleration speed) to be a combined 'representation' of the strength of the engine. The stronger the engine, the better the acceleration. The remaining variables: rear axle ratio (drat), the no of forward gears (gear) and weight (wt) are less or not related to the engine type and need to be taken into account. To determine which one of those variables play a influencing role, ANOVA is used to provide a variance table by fitting nested models with 'am' while adding respectively qsec, wt, drat and gear.

The code and resulting 'analysis of variance' table is provided in fig. 2 in the appendix. The resulting table shows that adding weight and qsec has a high significance (using the normal error level ($\alpha = 0.05$) and therefor should be added to the final model, but gear and drat do not add much and can be dropped.

## 4. Model fitting
The final model is a linear regression comparing transmission types on mpg (as outcome) while keeping weight en 1/4-mile time fixed on their average.

```{r model fitting, ECHO = TRUE}
fit <- lm(mpg ~ am + wt + qsec)   ## fitting the linear model
round(summary(fit)$coef,4)       ## looking at the results
```

The resulting table shows that the expected value for mpg is 9.62 for automatic transmission cars
with an average weigth and 1/4-mile time. Switching to manual transmission generates an increase of 2.94 mpg. Remarkable is also that the significance of the relationship between am and mpg, while taking wt and qsec into account is just slightly better than the normal error level ($\alpha = 0.05$). So in essence the hypothesis $H_0 = 0$ that claims that there is no relationship at all between am and mpg, is rejected.
The adjusted R-squared is 0.8336, so approx. 83% of the total variance is explained by the model used. The resulting confidence interval for the average increase going from automatic to manual transmission is 0.05 to 5.83 mpg (see figure 3 in the appendix). So with slightly more than 95% certainty we can predict that manual transmission will add mileage to the gallon.

## 5. Checking the residuals
Figure 3 in the appendix shows 4 residuals plots, to enable residual analysis. Residuals vs. Fitted is nicely scattered with no clear patterns and no hetero-skedasticity. The Normal Q-Q plot, appears to be pretty straight, showing that the residuals are mainly normally distributed. The Residuals vs. Leverage plot shows no points that go towards Cook's distance, indicating there are no real influential outliers.

## APPENDIX
### Figure 1. The correlation between all variables of mtcars

```{r CorOverview, echo=TRUE, message=FALSE, fig.align="center"}
g = ggpairs(mtcars, lower=list(continuous="smooth"))
g
```

### Figure 2. Analysis of variance table (ANOVA)

```{r anova, ECHO = TRUE}
fit1 <- lm(mpg~am)                         ## fitting just am
fit2 <- lm(mpg~am+wt)                      ## fitting am and weight
fit3 <- lm(mpg~am+wt+qsec)                 ## fitting am, weight and 1/4-mile time
fit4 <- lm(mpg~am+wt+qsec+drat)            ## fitting am, weight, 1/4-mile time and rear axle ratio
fit5 <- lm(mpg~am+wt+qsec+drat+gear)       ## fitting also no of gears
anova(fit1, fit2, fit3, fit4, fit5)
```

### Figure 3. The confidence interval for the increase of mpg for manual transmission

```{r ConfInterval, echo=TRUE, message=FALSE}
confint(fit)
```

### Figure 4. Residual-plots for fit

```{r ResidPlots, ECHO=TRUE, message=FALSE, fig.align="center"}
par(mfrow=c(2,2))    ## to plot 4 plots on one page
plot(fit)            ## plot the 4 standard residual plots
```
